---
title: "Proprioceptive precision in EDS"
author:
- Holly Clayton
- Bernard Marius 't Hart
- Denise Y.P. Henriques
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---


    • Aligned proprioceptive localization and reaching results.
    
    Figure 1a – baseline bias, still need; 
    Figure 1B is B on page 12 of R report (propEDS.docx) and is standard deviations of aligned localization. 
    Figure 1C is reach training endpoint errors. Is this the aligned data points in Figure B on page 5 of R report?; 
    Figure 1D – is the standard deviations of the training reach endpoint errors. I don’t think we have this.)
    
    • Learning curve results (I think maybe we chose to not include this since we have the aligned vs rotated aftereffect results in the next figure) (Figure 2 - Mention if there is difference in baseline cursor reaches; show without bootstrapped x. I don’t think we have this figure yet.)
    
    • Aftereffects results (Fig 3; use only without strategy rotated. We have Figure B on page 5 of the R report, but would have to delete “include” data points. I think we should include aligned results to show that they learned.)
    
    • Prop recal results (Figure 4; do we want all panels of figure on page 7 of R report?)
    
    • Prop/hypermobility correlation results (Figure 5 – figure on page 19 of R report)
    


# Introduction

In this study we want to replicate that people with Ehlers-Danlos Syndrome (EDS) have lower proprioceptive precision than age-matched controls [Clayton et al., XXXX]. We do this with a different method than before, which allows comparing proprioception with predicted sensory consequences. We also have Beighton scores for all participants so that we can see if hypermobility is correlated to proprioceptive precision. The paradigm is exactly the same as in a previous study [Modchalingam et al., 2019] from which we used the non-instructed variant with a rotation of 30 degrees. This could be used as a reference.

The main hypothesis is that in people with EDS, the proprioceptive signals are less reliable, so that precision of proprioceptive hand localization (without any other feedback) will also be less reliable.

The main measurements we use in this study come from a hand localization task. Participant's unseen, right hands are moved to some location. Directly above their right hand is a touch screen, and participants use the index finger of their visible, left hand to indicate where they think their right hand is. The movement with the right hand can be made by a robot, so that participants have an afferent, proprioceptive signal to estimate the location of their right hand. They can also self-generate the movement, so that in addition to afferent information, they also have efferent-based predicted sensory consequences to localize their hand with.

Participants do these two hand localization tasks after training movements with cursor that is aligned with the hand in a first session, and also after training with a cursor that represents the hand position, just rotated 30 degrees around the start position. Training with such rotated feedback is thought to lead to updates in efference-based predicted sensory consequences as well as proprioceptive recalibration. Both of these can be used in hand localization, but we think only the afferent, proprioceptive signal is affected by EDS.

Nevertheless, we will also analyze the data from the reach training session, to see if both people with EDS and controls learn at the same rate and to the same extent. While this is not directly relevant to our hypothesis, we will also test if implicit and explicit motor adaptation (as measured with a PDP procedure applied to reach aftereffects in no-cursor reaching). Here you can see all the data analyses we used, from pre-processing the raw data to published figures and reported statistics. All code and data can be checked and changed or even improved.

# Setting up the R environment

We recommend running the code in the R markdown notebook using R Studio, but it should be possible to run the functions from the R source files in a straightforward installation of R as well.

Packages we want:
```{r}
library('ez')
library('svglite')
```

We've kept the dependencies to a minimum. If either of the two packages is not available on your system, you can install them with this command: `install.packages(c('ez','svglite'))`.

There is some code we wrote to do all the pre-processing, create figures and do analyses. Importing the sources below will not do much by itself, as the files only contain functions we defined.
```{r}
source('R/common.R')
source('R/learningcurves.R')
source('R/reach_aftereffects.R')
source('R/localization_shifts.R')
source('R/localization_precision.R')
source('R/data.R')
```

If this ran without errors so far, all the functions we need are available.

We still need data. For this notebook, we can simply download pre-processed data files from OSF, but you can choose to download the raw data, and run / check / change pre-processing functions yourself. To do that change the variable `usePreprocessed` to FALSE.

```{r}

usePreprocessed <- TRUE
#usePreprocessed <- FALSE

if (usePreprocessed) {
  
  # we make sure the files are downloaded from OSF and unzipped:
  getPreprocessedData()
  
} else {
  
  # we still need to download the raw data from OSF:
  getPreprocessedData(participantsOnly=TRUE)
  
  # and then we run a bunch of function to get out the interesting bits:
  
  # This function is from data.R and downloads and unzips a zip file with the raw data:
  getRawData()
  
  # These functions are from learningcurves.R and 
  # 1) create the pre-processed learning-curve files:
  getLearningCurves()
  # 2) create files with reach precision during training:
  getReachPrecision()
  
  # This function is from reach_aftereffects.R creates
  # 1) the nocursors / reachaftereffect files, and
  # 2) the no-cursor precision files:
  getAllNoCursors()
  
  
  # The following functions are from localization_shifts.R.
  # This one creates the corrected [group_]localization.csv files:
  getLocalization()
  # This one calculates 95% confidence intervals for plotting:
  getLocalizationTdistributionConfidenceIntervals()
  # This one creates the loc_AOV files to do statistics on:
  getLocalizationForANOVA()
  
  # These functions are from localization_precision.R and
  # 1) create the files with participants' localization variance:
  getLocalizationVariance()
  # 2) create the files with participants' localization biases:
  getLocalizationTrends()
  
}
```

Now that both code and data are available, we can start.

_Planned Figures_

For the Methods:

1. Setup & Trial types
2. Task order

For the Results:

3. Baseline proprioceptive accuracy & precision
4. Aligned reaching
5. Learning curves
6. No-cursor data
7. Recalibrated proprioceptive accuracy & precision
8. Beighton scores over proprioceptive precision

_Current Figures_

For the Methods:

1. Setup & Trial types
2. Task order

For the Results:

3. Reach training, with aligned and rotated reach precision
4. Implicit no-cursor changes, with aligned and rotated open-loop reach precision
5. Localization: baseline biases, shifts, precision
6. Beighton scores over proprioceptive precision

- 3 & 4 show no difference between EDS and controls
- 5 does show an effect on precision (which can't be explained by closed/open-loop baseline, changes or precision: see 3+4)
- 6 and this correlates with severity of symptoms

# Proprioceptive accuracy and precision

> _Our main question is on the proprioceptive accuracy and precision in people with EDS compared to controls. However, before we able to investigate this with our data, we need to rule out a few alternative explanations. First, do people with EDS reach differently to begin with, or adapt in a different way compared to controls. Second, is the implicit motor adaptation in people with EDS different from controls?_

# Adaptation in people with EDS and Controls

First, we look at the adaptation to the 30 degree rotation, to see if the two groups adapt an equally amount and equally fast.

Plot learning curves:
```{r fig.width=8, fig.height=6}
plotLearningCurves(target='inline')
```

So it looks like the people with EDS and the controls learn about as much and equally fast. But we'll put that to the test here.

Learning curves ANOVA:
```{r}
learningCurveANOVA()
```

This ANOVA finds no difference between the two groups, nor an interaction between group and block. There is an effect of block. We interpret this to mean that both groups learned, and that there is no discernable difference in either their rate of learning or in their asymptotic level of learning.

## Closed-loop reach precision

We also see the reach precision (the standard deviation of target-normalized reach errors, for the last 15 trials of the first long block in both the aligned and rotated session). There might be a small effect of group as the control group seems to have higher precision in the aligned session, but their precision decreases more than that of the EDS group. We this this with an mixed-design ANOVA on reach precision using group (control or EDS) as a between-subjects factor and training (aligned or rotated) as a within-subjects factor.

```{r}
reachPrecisionANOVA()
```

There is an effect of training session (aligned or rotated), so that we can say that training with manipulated feedback decreases precision. However, there is no effect of group (EDS or control), nor an interaction between group and training session. We take this to mean that the precision of closed-loop reaches is comparable between EDS and control groups, and also changes comparably with rotated training.

> _There is no difference between the EDS and control group in how well or how quickly they adapted to the rotation, nor in their baseline closed-loop reach precision or it's decrease in the rotated session. That is: there are no a priori differences in closed-loop reaching between the groups that could be alternative explanations for performance differences on other tasks._

# Reach Aftereffects in people with EDS and Controls

First a figure:
```{r fig.width=5.5, fig.height=5.5}
plotReachAftereffects(target='inline')
```

Now, we'll compare the aligned no-cursor deviations with no-cursor deviations after rotated training, in a mixed-design ANOVA with the between subjects factor _group_ (EDS or control) and the within subject factor _training_ (aligned or rotated).

```{r}
NoCursorANOVA()
```

There is an effect of training, but no effect of group and no interaction between group and training. We interpret this to mean that there is no discernable difference between the groups in their level of implicit adaptation as measured by the shift from no-cursor reach deviations after aligned cursor training to those in "without strategy" no-cursor reach deviations after rotated training.

We do not analyse the ability to include strategies in reach aftereffects. We expect neither the controls nor the EDS group to be able to do this, as they receieved no instructions (Modchalingam et al., 2019). However, there are some functions in 'R/reachaftereffects.R' which show that unexpectedly, the controls do show a difference between with and without strategy reach aftereffects, while the EDS group does not. This seems to mean that the controls here behave atypically, but it is beside the point. Here we included the with strategy and without strategy conditions to make sure that in the without strategy condition, we really have a measure of implicit adaptation, which is what we are really interested in. While we don't know how the control group here did acquire a strategy, at least the EDS group is indistinguishable from the participants doing an identical protocol in our earlier study, and implicit learning is equal in both our EDS and control group.

## No-cursor reach precision

Closer to our interest is the precision with which people reach while not seeing their hand. First, we'll test if there is an a-priori difference between the two groups, by comparing the precision in the aligned condition, using a t-test.

```{r}
RAEprecisionTtest()
```

There is no a priori difference in no-cursor reach precision between the groups.

To test if the rotated training induced different changes in the precision for the two groups, we'll do a mixed-effect ANOVA on no-cursor reach precision using training (aligned or rotated) as a within-subjects factor and group (control or EDS) as a between-subjects factor.

```{r}
NoCursorANOVA(datatype='precision')
```

As with reach aftereffects, there is an effect of training, but there is no effect of group or interaction between group and training. This ANOVA and the previous, taken together mean that the reach planning (internal models, if we must) is updated in an equal way in both groups.

> _There is no difference between the EDS and control group in the extent of implicit reach aftereffects, or the precision of open-loop reaches and how they change with rotated training. This means these are ruled out as alternative explanations for any other differences we find in localization._

# Localization

So now we will look at the measure of interest: hand localization. But before testing the precision of hand localization, we will first test if it shifts and if it shifts differently from the control group. Based on our earlier study with people with EDS, we do not expect this.

First a plot of the shifts in localization responses, induced by rotated training:

```{r fig.width=8, fig.height=8}
plotLocalizationShifts(target='inline')
```

To clarify, all data have the aligned localization responses subtracted, except the first two panels, which show the interpolated trends in the aligned localization tasks.

## Baseline biases

**No analyses done yet.** Looks like there is no difference.

## Training induced localization shifts

Looks like there is an effect of training on localization (data in panels C & D seem different from 0). It also looks like there is no difference between the two groups (controls and people with EDS) in both the active and passive localization. It does seem like there is a difference between shifts in active and passive hand localization (data in panel E is somewhat different from 0), but it's unclear if there is a group difference there. Let's run a bunch of ANOVA's.

What are the factors?

angle = hand angle: one of 45, 90 and 135

```{r}
localizationAOV()
```

This mostly confirms what the figure shows: effects of training on localization, a difference between active and passive localization, but no effects of group.

> _There is no difference between the EDS and control group in their localization shifts, so this can not be an alternative explanation for any differences we might find in localization precision._

## Localization precision

Now the interesting part. We've mostly not seen any differences between the two groups which is what we expected. The exception is the _with strategy_ reach aftereffects, but for the current study, the more important issue is that there were no group effects on any of the localization measures. The question is if using this fast method of assessing proprioceptive recalibration we find the same effect as earlier (Clayton et al., XXXX) [@Clayton1234]: that people with EDS have larger/smaller proprioceptive precision. (The wider confidence interval in the group of people with EDS in the difference between active and passive localization shifts may be due to differences between individuals localization shifts, but not their proprioceptive hand localization precision.)

After subtracting a smoothed spline fit to each individual participants localization errors across hand angles, in each of the 4 conditions, which should factor out individual levels of accuracy, we are then left with the precision of their responses in all 4 conditions. Of those we take the variance, but for frequentist stats, it is more useful to use the standard deviation (as it gets the data more or less normally distributed), so we convert the variance to standard deviation before doing statistics.

In the figure above, we can also see the standard deviations (square root of the variance) as our measure of precision of localization, for both the control group (in red) and the people with EDS (in blue). In panel A we see means and an estimate confidence interval assuming the data comes from a sample t-distribution. On the right is the localization precision in the aligned session, on the right in the rotated session. The continuous line is for active localization and the dashed line for passive. In panels B and C the individual data and the bootstrapped confidence interval are shown. Panel B is for the aligned session, and panel C for the rotated session. Each shows active (left) and passive (right) localization precision, for both the control group (red) and the group of participants with EDS (blue). 

In the figure it appears that localization is less precise in the group of people with EDS as compared with the control group.

Let's test this. Below is a mixed-design ANOVA on the standard deviation of localization, when using the group (control or EDS) as between-subjects factor and passive/active as well as aligned/rotated as within-subject factors.

```{r}
localizationSTD.ANOVA()
```

There is a main effect of group as well as a main effect of passive versus active localization, but there are no higher order effects. Let's do some one-sided t-tests to look into our hypothesis. One-sided, because we think that people with EDS should be less precise in localizing their hand:

```{r}
localicalizationSTDt.tests()
```

So apart from in aligned & active localization, people with EDS are indeed less precise in localizing their hand as compared to age-matched controls without EDS.

It also seems that the difference between active and passive is mostly the same within all conditions and groups. But, as mentioned, for the EDS group, active localization after aligned training might be more precise (lower variance) than in the other three conditions. Perhaps before changing it up with adaptation, the signal from predicted sensory consequences provides enough extra information to make a difference for hand location estimation.

> _As we predicted there is a difference in the precision of localization between people with EDS and controls: people with EDS seems to have slighthly lower hand localization precision. Importantly, this can not be explained by closed-loop reach precision, adaptation rate and extent, open-loop reach precision, implicit adaptation and hand localization shifts. Interestingly, there is no such difference in the aligned, active localization._

# EDS indicators and hand localization precision

So if the precision of limb localization is lower in the EDS group, then perhaps there is a relationship between EDS indicators and the precision of limb localization. We use Beighton scores as a hypermobility index, measured in both groups, and test if this is correlated with overall localization variance (variance is now calculated across all 4 conditions for every participant - providing more power). First, let's plot this data, with a linear regression:

```{r fig.width=4, fig.height=4}
plotBeightonLocSTD(target='inline')
```

This looks like there is some relationship between the two variables, and this is confirmed by a Pearson correlation:

```{r}
correlateBeightonLocSTD()
```

# Predict EDS from hand localization precision

This raises the question of whether or not EDS can be predicted from hand localization variance. The figure above shows there are quite a few people with EDS that are in the range of the control group though, so this might not be the case. However, with the localization variance calculated across all 4 conditions it does work to some extent:

```{r}
predictGroupBYlocvar()
```

Although I'm not using a training/test data split here which would be more appropriate, and perhaps we can make use of some other predictors as well (e.g. train a random forest that can make use of localization variance, and compare it to one that can not use localization variance), but all of this would require a lot more data (participants).

We can explore one issue with such a test: it would be best if it were cheap / easy to do. Training on a visuomotor rotation takes time, so the data from the rotated session might not be useful. Additionally, the passive localization requqires an expensive piece of equipment, which may be prohibitive. Therefore we should test how well predicting the group (people with EDS vs. controls) from localization variance works within each of the four hand localization conditions:

```{r}
localizationSTDlogreg()
```

Unfortunately, the predictions are useless in the condition that would be most easily used in a diagnostic setting. This is still done without training/test data splits, and still ignores co-predictors, which could reduce and increase power respectively. Nevertheless, at first glance, it seems that using hand localization precision is not a usefull predictor of EDS.

# Summary

Here is a list of findings / results:

- The EDS and control group adapted equally.
- Closed-loop reach precision, as well as the increase in closed-loop reach precision with rotated training was comparable between groups.
- The EDS and control group had equal implicit (without strategy) reach aftereffects.
- EDS and control group had comparable open-loop reach precision as well as a comparable rotation-induced increase in open-loop reach precision.
- The EDS and control group shifted their localization responses equally, and had comparable aligned localization errors overall.

This rules out many alternative explanations for the interesting findings:

- The EDS group was less precise in hand localization compared to the control group, except perhaps in the aligned/active condition (as predicted / confirming our earlier paper).
- Beighton scores predict overall hand localization precision (across both groups), and while hand localization precision has some predictive power of which group a participant is in, it's differentiation is not great.

# Store figures

This chunk of code merely stores all the figures as SVG files in the `doc/` folder of the project:

```{r}
plotLearningCurves(target='svg')
plotReachAftereffects(target='svg')
plotLocalizationShifts(target='svg')
#plotLocalizationVariance(target='svg')
plotBeightonLocSTD(target='svg')
```

